---
title: "QGM Assessment 2025"
author:
date: "23/01/2024"
output: 
  html_document:
    number_sections: TRUE
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
options(width=128)
knitr::opts_chunk$set(echo = TRUE)
```

The asreml library and the qgm helper functions are loaded below

```{r}
setwd("~/Semester 2/Quantitative Genetic Models/Assessment P1")
library("asreml")
source("qgm_helpers.R")
```

```{r}
tut4 <- read.csv(file="QGM_dat_2025_P1.csv", header=TRUE, na.strings=c("NA"))
head(tut4)
tail(tut4)
```

```{r}
tut4[c(2001,2701),]
tut4[c(1754,1748,1089,1738),]
tut4[c(913,733,574,323,49,574),]
```

```{r}
tut4[c(1,1001),]
```
```{r}
gen0_data <- tut4[tut4$generation == 0, "dbh21"]
shapiro.test(gen0_data)

gen1_data <- tut4[tut4$generation == 1, "dbh21"]
shapiro.test(gen1_data)
```

```{r}
str(tut4$dbh21)
mean(tut4$dbh21, na.rm = TRUE)
sd(tut4$dbh21, na.rm = TRUE)
```

```{r}
sires1 <- sort(unique(tut4$sid))
dams1 <- sort(unique(tut4$did))
dams1 ==  sires1
dams1
```


```{r}
mean(tut4$dbh21[tut4$generation == 0], na.rm = TRUE)
mean(tut4$dbh21[tut4$generation == 1], na.rm = TRUE)
sd(tut4$dbh21[tut4$generation == 0], na.rm = TRUE)
sd(tut4$dbh21[tut4$generation == 1], na.rm = TRUE)
```

```{r}
addmargins(table(tut4$gen,tut4$treatment))
```
Control: treatment = 1
Experimental root treatments: treatment = 2/3
Generations: 0-2
**Have no phenotypic data on generation 2**

```{r}
contingency_table <- table(tut4$sid, tut4$did)
contingency_df <- as.data.frame(contingency_table)
filtered_df <- contingency_df[contingency_df$Freq != 0, ]
print(filtered_df)
```

```{r}
library(ggplot2)

# Create a summary table
summary_table <- as.data.frame(table(tut4$generation, tut4$treatment))
colnames(summary_table) <- c("Generation", "Treatment", "Count")

# Plot the data
ggplot(summary_table, aes(x = Generation, y = Count, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Treatments per generation",
       x = "Generation",
       y = "Count") +
  scale_fill_discrete(name = "Treatment") +
  theme_minimal()
```
Unbalanced

```{r}
qgm_ped_summary(tut4[,1:3])
```
300O ids, only 1 id/individual
1000 have no sire or dam
100 distinct sires, 20 offspring each
100 distinct dams, 20 offspring each
Each breeding individual produces 40 offspring
100 mating pairs, only 1 dam/sire each
**100 parents used as sire and dam: listed**

```{r}
library(ggplot2)

# Create the data frame
pedigree_data <- data.frame(
  Category = c(
    'No. of distinct ids',
    'No. of distinct ids with no sire or dam',
    'No. of distinct sires', 
    'No. of distinct dams', 
    'No. of mating pairs', 
    'No. of sires per dam', 
    'No. of dams per sire', 
    'No. of offspring per pair', 
    'No. of parents used as sires and dams'),
  Count = c(3000, 1000, 100, 100, 100, 1, 1, 20, 100)
)

# Create the bar plot
ggplot(pedigree_data, aes(y = Count, x = reorder(Category, Count))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Summary of Pedigree Information", y = "Count", x = "Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = Count), vjust = -0.3, size = 3.5) +
  ylim(0, 3300)
```


```{r}
tut4$treatment <- as.factor(tut4$treatment)
tut4$generation <- as.factor(tut4$generation)
tut4$tid <- factor(tut4$tid) 
tut4$sid <- factor(tut4$sid)
tut4$did <- factor(tut4$did)
```

```{r}
hist(tut4$dbh21)
points(tut4$generation)
```
```{r}
ggplot(tut4, aes(x = dbh21, fill = factor(generation))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  labs(x = "DBH21", y = "Frequency", fill = "Generation") +
  theme_minimal()
```


```{r}
library(ggplot2)
ggplot(tut4, aes(x = treatment, y = dbh21, color = treatment)) +
  geom_point(size = 3) +
  labs(title = "DBH21 by Treatment",
       x = "Treatment",
       y = "DBH21") +
  scale_color_discrete(name = "Treatment") +
  theme_minimal()
```

```{r}
install.packages("gridExtra")
library(gridExtra)
library(ggplot2)

# Histogram plot
p1 <- ggplot(tut4, aes(x = dbh21, fill = factor(generation))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  labs(title = "DBH21 by Generation", x = "DBH21", y = "Frequency", fill = "Generation") +
  theme_minimal()

# Scatter plot
p2 <- ggplot(tut4, aes(x = treatment, y = dbh21, color = treatment)) +
  geom_point(size = 1.5) +
  labs(title = "DBH21 by Treatment",
       x = "Treatment",
       y = "DBH21") +
  scale_color_discrete(name = "Treatment") +
  theme_minimal()
grid.arrange(p1, p2, ncol = 2)
```


```{r}
library(ggplot2)

ggplot(tut4, aes(x = treatment, y = dbh21, color = treatment)) +
  geom_point(size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black", aes(group = 1)) +
  labs(title = "DBH21 by Treatment",
       x = "Treatment",
       y = "Diameter") +
  scale_color_discrete(name = "Treatment") +
  theme_minimal()
```


```{r}
qgm_ped_summary(tut4[tut4$gen %in% c(0),])
```

```{r}
qgm_ped_summary(tut4[tut4$gen %in% c(1),])
```

```{r}
qgm_ped_summary(tut4[tut4$gen %in% c(2),])
```
All as expected

```{r}
ainv <- ainverse(tut4[,1:3]) 
```


```{r}
tut4.asr <- asreml(fixed=dbh21~1+treatment,
              random=~vm(tid,ainv),
              residual=~idv(units),
              data=tut4)
```
y = 1m + Xb + Zu + e

y = the (2000 x 1) vector for dbh21 in the trees.

1 = the (2000 x 1) vector of 1s, the identity matrix.

m = a fitted mean, scalar.

b = the (2 x 1) vector of fixed effects with 2 d.f. for treatment

X = the (2000 x 2) design matrix for fixed effects.

u = the (3000 x 1) vector of BREEDING VALUES assumed to be RANDOM with DISTRIBUTION MVN(0,VaA), where A is the NUMERATOR RELATIONSHIP MATRIX and Va is the ADDITIVE GENETIC VARIANCE.

Z = the (2000 x 3000) design matrix for random effects.

e = the (2000 x 1) vector of residuals assumed to be random and distributed MVN(0,VeI) where I is the (2000 x 2000) Identity matrix and Ve is the residual variance). 

```{r}
residuals <- residuals(tut4.asr)
residuals_df <- data.frame(Residuals = residuals, Fitted = fitted(tut4.asr))
library(ggplot2)

ggplot(residuals_df, aes(x = Fitted, y = Residuals)) +
  geom_point(size = 2, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

```{r}
# Extract residuals from the asreml object
residuals <- residuals(tut4.asr)

# Create the Q-Q plot
qqnorm(residuals, main="Normal Q-Q Plot", 
       xlab="Theoretical Quantiles", 
       ylab="Sample Quantiles", 
       pch=19, col="blue")

# Add a reference line
qqline(residuals, col="red", lwd=2)

# Enhance the plot with additional styling
grid()  # Add a grid for better readability
```

```{r}
# Control vs. Treatment 2
ct1 <- c(1, -1, 0)
qgm_fixed_linear(tut4.asr, "treatment", ct1)
```
Control mean ~0.1 less

```{r}
# Control vs. Treatment 3
ct2 <- c(1, 0, -1)
qgm_fixed_linear(tut4.asr, "treatment", ct2)
```
Control mean ~0.5 less

```{r}
# Treatment 2 vs. Treatment 3
ct3 <- c(0, 1, -1)
qgm_fixed_linear(tut4.asr, "treatment", ct3)
```
Treatment 2 ~0.4 less that 3

```{r}
# Control vs. Average of Treatments
ct4 <- c(1, -0.5, -0.5)
qgm_fixed_linear(tut4.asr, "treatment", ct4)
```
Control mean ~0.3 less that treatments averages

```{r}
tut6.asr <- asreml(fixed = dbh21 ~ 1 + treatment,
                   random = ~ vm(tid, ainv)+idv(jid),
                   residual = ~idv(units),
                   data = tut4,
                   as.is = TRUE)

# Summary of the model
qgm_summary(tut6.asr)
```

```{r}
vpredict(tut6.asr, h2 ~ V2 / (V1 + V2 + V3))
```


```{r}
# Remove rows with missing values
tut4_clean <- na.omit(tut4)

# Refit the model with the cleaned data
tut8.asr <- asreml(fixed = dbh21 ~ 1 + treatment,
                   random = ~ vm(tid, ainv) + idv(jid) + idv(did),
                   residual = ~idv(units),
                   data = tut4_clean,
                   as.is = TRUE)

# Summary of the model
summary(tut8.asr)
#does not work
```

```{r}
vpredict(tut8.asr, h2 ~ V3 / (V1+V2+V3+V4))
```

```{r}
qgm_summary(tut4.asr)
```

Additive genetic variance = 12.867742 with se 1.3397653
Residual variance = 5.182249 with se 0.8874169

```{r}
p_value_treatment_2 <- 2 * (1 - pnorm(abs(0.3758349)))
p_value_treatment_3 <- 2 * (1 - pnorm(abs(1.8640447)))

p_value_treatment_2
p_value_treatment_3
```


```{r}
print(as.data.frame(wald(tut4.asr,denDF="numeric",ssType="conditional")$Wald)[,c(1,2,4,6)])
```
At least 1 treatment differs from the others.

```{r}
vpredict(tut4.asr, phenotypic_variance ~ V1 + V2)
```

```{r}
vpredict(tut4.asr, h2 ~ V1 / (V1 + V2))
```

```{r}
sqrt(0.7128946)
sqrt(0.7539588)
```

```{r}
head(qgm_random_effects(tut4.asr, "tid"))
ebv <- qgm_random_effects(tut4.asr,"tid")
head(ebv)
```
```{r}
nrow(ebv)
colnames(ebv)[1] <- "ebv"
colnames(ebv)[2] <- "se"
colnames(ebv)[4] <- "tid"
head(ebv)
```
```{r}
pedx <- merge(tut4, ebv, by = "tid", sort = FALSE)
head(pedx)
tail(pedx)
nrow(pedx)
```

```{r}
par(mfrow = c(1, 2))
plot(pedx$generation, pedx$ebv, col="blue", cex=0.1, xlab="Generation", ylab="Estimated Breeding Value")
points(pedx$generation, pedx$ebv, col="red", cex=0.1)
plot(pedx$treatment, pedx$ebv, col="purple", cex=0.1, xlab="Treatment", ylab="Estimated Breeding Value")
points(pedx$treatment, pedx$ebv, col="pink", cex=0.1)
par(mfrow = c(1, 1))
```
No selection?

```{r}
pedx$f <- attr(ainv,"inbreeding")
bva <- 12.867742
pedx$b_acc <- sqrt(1. - pedx$se**2./(bva*(1+pedx$f)))
head(pedx$b_acc)
head(pedx)
tail(pedx)
#Additive genetic variance component
```

```{r}
mean(pedx[pedx$gen %in% c(0),"b_acc"])
mean(pedx[pedx$gen %in% c(1),"b_acc"])
mean(pedx[pedx$gen %in% c(2),"b_acc"])
```
```{r}
mean(pedx$b_acc)
```


```{r}
pedx[pedx$tid %in% c(1,49,1001,1057,2001,2701), "b_acc"]
```
```{r}
plot(pedx$b_acc, pedx$gen, ylim = c(0,4), xlim = c(0.4,1.05))
```

```{r}
plot(pedx$b_acc, pedx$gen, ylim = c(0, 4), xlim = c(0.4, 1.05), 
     xlab="Breeding Accuracy", ylab="Generation", 
     main="Breeding Accuracy vs Generation", 
     pch=19, col="blue")

# Adding grid lines
grid()
```
Accuracy gets worse across the generations. 

```{r}
unique(pedx$b_acc)
pedx$b_acc
```

```{r}
par(mfrow = c(1, 2))

# First plot: Treatment vs. Accuracy
plot(pedx$treatment, pedx$b_acc, col = "darkblue", pch = 16, cex = 0.5, xlab = "Treatment", ylab = "Accuracy", main = "Accuracy vs Treatment")
points(pedx$treatment, pedx$b_acc, col = "darkblue", pch = 16, cex = 0.5)

# Second plot: Generation vs. Accuracy
plot(pedx$generation, pedx$b_acc, col = "darkblue", pch = 16, cex = 0.5, xlab = "Generation", ylab = "Accuracy", main = "Accuracy vs Generation")
points(pedx$generation, pedx$b_acc, col = "darkblue", pch = 16, cex = 0.5)

# Reset the plotting area to default (optional)
par(mfrow = c(1, 1))
```


```{r}
par(mfrow = c(1, 2))
hist(pedx[pedx$generation %in% c(0),]$ebv, breaks = 20, main = "EBV Histogram", xlab = "EBV")
hist(pedx[pedx$generation %in% c(0),]$se, breaks = 20, main = "SE Histogram", xlab = "SE")
par(mfrow = c(1, 1))
```
```{r}
par(mfrow = c(1, 2))
hist(pedx[pedx$generation %in% c(1),]$ebv, breaks = 20, main = "EBV Histogram", xlab = "EBV")
hist(pedx[pedx$generation %in% c(1),]$se, breaks = 20, main = "SE Histogram", xlab = "SE")
par(mfrow = c(1, 1))
```
```{r}
par(mfrow = c(1, 2))
hist(pedx[pedx$generation %in% c(2),]$ebv, breaks = 20, main = "EBV Histogram", xlab = "EBV")
hist(pedx[pedx$generation %in% c(2),]$se, breaks = 20, main = "SE Histogram", xlab = "SE")
par(mfrow = c(1, 1))
```
Bigger spread as predictions get better. 

```{r}
pedx[pedx$tid %in% c(1,49,1001,1057,2001,2701),]
```
```{r}
pedx[pedx$tid %in% c(2001,2002,2003),]
```
Same info, have the same parents


```{r}
pedx[pedx$tid %in% c(1754,1089),]
```
Average of the parents breeding values

```{r}
print(as.data.frame(summary(tut4.asr,coef=TRUE)$coef.fixed))
```
(Intercept)	29.6790296	0.2685880	110.5002242	
treatment_1	0.0000000	NA	NA	
treatment_2	0.1052886	0.2801460	0.3758349	
treatment_3	0.5192144	0.2785418	1.8640447		


```{r}
tut5.asr <- asreml(fixed=dbh21~1+treatment,
                   residual=~idv(units),
                   data=tut4)
qgm_summary(tut5.asr)
```
#2x(logL1-logL0)
```{r}
2 * (-3707.27086944953 - -3861.36553448811)
qchisq(0.95,1) #type 1 error = 0.05
```
308.1893 >> 2.705543

```{r}
qchisq(0.998,1)
```

```{r}
tut7.asr <- asreml(fixed=log(dbh21)~1+treatment,
                   random=~vm(tid,ainv),
                   residual=~idv(units),
                   data=tut4,
                   as.is = TRUE)
qgm_summary(tut7.asr)
```
```{r}
plot(tut7.asr)
```

```{r}
vpredict(tut7.asr, h2 ~ V1 / (V1+V2))
```

```{r}
# Extract the estimated breeding values (assuming they are stored in a column named 'ebv')
ebvs <- pedx$ebv

# Calculate the correlation
correlation <- cor(ebvs, residuals, use = "complete.obs")

# Print the correlation
print(correlation)
```
```{r}
residuals2 <- residuals(tut7.asr)
cor(ebvs, residuals2, use = "complete.obs")
```
```{r}
shapiro.test(residuals)
```
```{r}
head(pedx)
pedx[0:10,]
min(pedx$ebv)
tail(pedx)
```

```{r}
mean(pedx$ebv)
```


